---
title       : Market Basket Analysis
subtitle    : Recommender System in R
author      : Dhiraj Khanna
job         : 
framework   : io2012        # {io2012, html5slides, shower, dzslides, ...}
highlighter : highlight.js  # {highlight.js, prettify, highlight}
hitheme     : highlight      # 
widgets     : [mathjax]            # {mathjax, quiz, bootstrap}
mode        : selfcontained # {standalone, draft}
knit        : slidify::knit2slides

---
<style>
em {
  font-style: italic
}
strong {
  font-weight: bold;
}
</style>

## Introduction to Market Basket Analysis (MBA)

1. Looks at purchase coincidence
2. It investigates whether two products are being purchased together, and whether the purchase of one product increases the likelihood of purchasing the other
3. Data arranged in large matrix of rows and columns
4. Rows represent **shoppers** and columns are different **products**
5. Entries in the matrix are 1 or 0.
6. MBA used by last.fm and Netflix

--- .class #id 

## The Math behind MBA

- Look for **association rules** such as $A = LHS (\text{buy chips}) \to B = RHS (\text{buy beer})$
- LHS is **antecedent**, RHS is **consequent** and arrow is **is related to**
- "Confidence" is the condidtional probability of RHS, given LHS $$P(B|A) = \frac{P(A \text{and} B)}{P(A)}$$
- **Support** is the frquency of A or B and is calculated as $P(A)$ or $P(B)$
- *Lift of A on B* $$ lift(A \to B) = \frac{P(B|A)}{P(B)}$$
- If lift > 1, things are good
- But big lifts are relevant only if the confidence is also high
- *Leverage* of A = LHS on B = RHS as $$leverage(A \to B) = P(A \text{and} B) - P(A)P(B)$$

---
## MBA on Last.fm dataset

Suppose you were given data from a music community site. For each user you
may have a log of every artist he/she had downloaded to their computer. You
may even have demographic information on the user (such as age, sex, location,
occupation, and interests). Your objective is to build a system that recommends
new music to users in this community.

---

## Analysis in R

Let's read in the data

```{r}
lastfm <- read.csv("data/lastfm.csv")
head(lastfm)
```

---
Convert "user" into a factor

```{r}
lastfm$user <- factor(lastfm$user)
```

Now we will use the base `split()` function to split the artists list by "user" and remove duplicate entries of artists. It is important to do so becuase later when we use the `arules` package, it is necessary to have unique values as it is not possible to coerce list with transactions with duplicated items.

```{r}
playlist <- split(x=lastfm[,"artist"],f=lastfm$user)
playlist[1:2]
```

---

An artist may be mentioned by the same user more than once, hence it is important to remove artist duplicates before creating the incidence matrix.

```{r}
playlist <- lapply(playlist,unique) ## remove artist duplicates
```

---

We now have a list of users and the artists (unique values) they listen to. Now we will use the `arules` package for the rest of the assignmnet.

```{r, message=FALSE}
library(arules)
playlist <- as(playlist,"transactions")
```
This converts the playlist to a data class defined in `arules` package as transactions. We can now set a threshold for the required support.

---

```{r}
itemFrequencyPlot(playlist,support=.08,cex.names=1.5)
```

---
To look at the top three most popular artists :
```{r}
head(sort(itemFrequency(playlist), decreasing=T),3)
```

---
Now let us create the association rules using the `apriori` function with a support of 0.01 and confidence of 0.45.

```{r}
musicrules <- apriori(playlist,parameter=list(support=.01, confidence=.45))
```
---
Now on these rules, we will make a subset of only those rules which provide a lift of greater than 6.

```{r}
inspect(sort(subset(musicrules, subset=lift >= 6), by="confidence"))
```

---
Again heartening to see that classic rock still figures in the top 3 as far as confidence is concerned. We could convert all of this to a data frame and write it as a csv file.

```{r}
lift6 <- sort(subset(musicrules, subset=lift >= 6), by="confidence")
lift6 <- as(lift6, "data.frame")
write.csv(lift6, "lift6.csv")
```

---
The top 3 rules having the highest confidence are:
```{r}
head(lift6, 3)
```

---
We will use the `arulesViz` library for plotting all the rules with a support of 0.01 and confidence of 0.45.

```{r, message=FALSE}
library(arulesViz)
plot(musicrules, method="grouped", interactive=T)
```